Ancestry linkage in hybrid zones

### 2016-02-16 drafts/hybrid_zone.tex
fixed some references

Within a genome, the mixing of ancestries is brought about by the the process of recombination that takes place every generation \yb{(CITE Fisher1954, Chapman \& Thomson; Barid et al)}.   
The expected distribution of lengths of unbroken ancestry have been previously described under neutral conditions \cite[e.g.][]{Gravel2012,Sedghifar2015}.  
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection. 
Selection against migrant \yb{alleles} in a hybrid zone results in reduced frequencies of foreign ancestry surrounding targets of selection \yb{(CITE CLINES)}. 
The effect on patterns of ancestral block lengths is, however, less well understood. 
Because of rapid removal from the population, we expect to observe \yb{a deficit of} short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. 
\alisa{While empirical patterns of block length in hybrid zones have not been well studied, the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!

@@

\section{Methods}
\subsection*{The Model}
We consider two isolated populations --- labeled \emph{Species $A$} and \emph{Species $B$} ---  that come into contact $\tau$ generations in the past. 
Individuals move forward in time by sampling a displacement from a Gaussian distribution. 
We consider a simple case of under-dominance. 
\yb{Although most selection in hybrid zones is likely more complex than the single locus intrinsic under-dominance we consider, previous work has shown that such models generate clines that are effectively indistinguishable from more complex models of ecological selection \citep{Barton1989, BARTON and GALE 1993} which also generate clines like those degenerated under complex epistatic systems}. Specifically, here the fitness of an individual is  determined by its genotype at a particular locus $L$, such that individuals that are heterozygous at locus $L$ have fitness reduced by $s>0$ compared to homozygotes. To fix notation, we will say that species $A$ was initially on the ``left'' of the zone of contact, which corresponds to spatial positions $x<0$.

@@

\paragraph{Establishment of  \sout{the selected cline} \alisa{a cline at the selected locus}}
Consider a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline.
To study gene flow past this cline we first need to understand how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring.
Assuming local random assortment of alleles into diploids,
that the habitat and dispersal are homogeneous,
and that $s$ and $\sigma$ are small, 
the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{bazykin}; for discussion of diploidy see \citep{diploidcline}.
There is not an exact analytic solution to this equation, 
but a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{bazykin}.
Considering the numerous approximations going into this,
the important attributes are that 
the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially.
Furthermore, the cline is established through diffusion \alisa{of individuals?}, which is slowed down somewhat by selection against heterozygotes,
but is still diffusive, and so it will take time of order $1/s$
to reach its stable width \alisa{(is this something that has been shown before?)} This can be shown by rescaling space by $\sigma/s$ and time by $1/s$ in equation \eqref{eqn:cline_pde}, and seeing that the result \alisa{(what result?)} doesn't depend on $\sigma$ or on $s$.

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is given.
We will understand how lineages of sampled individuals behave,
both at and away from the selected site,
by following lineages as they move between backgrounds,
as done for instance in \citet{selectioncoal,ralph2015patchy}.
Consider the collection of $A$ alleles found at location $x$.
The expected number of such alleles produced by location $y$ is proportional to 
the number of $A$ alleles at $y$ multiplied by the probability of dispersing from $y$ to $x$;
looking in the other direction, this is also the probability that an $A$ allele at $x$ had a parent at $y$.
In other words, lineages move as the random walk determined by dispersal,
but biased towards areas that produce more of the selected allele that they carry.
Making the same assumptions underlying equation \eqref{eqn:cline_pde},
we show in Appendix \ref{apx:lineage_derivation}
that the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $s \grad \log p(x,t)$ (i.e., it is pushed uphill on $\log p$ at speed $s$) \alisa{note to self to read through appendix when written}.
Since $p=1$ far from the zone to the left, 
and is proportional to $\exp(-x\sqrt{s}/\sigma)$ to the right,
then roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma/\sqrt{s}$ 
back towards the side where they are more common.
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact,
this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction.

@@

\paragraph{How fast do linked clines relax?}
Although loci not under selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{bartonbarrier}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
Suppose we sample across at cline $\tau$ units of time after secondary contact.
How far away from the selected locus do we expect to see clines that have begun to flatten out more
than the selected cline?
A linked cline will have begun to flatten out if there is an appreciable chance that
a lineage at the linked locus initially linked to an $A$ allele 
at some point has recombined from an individual carrying the $B$ allele.
This implies that loci closer than $1/\tau$ to the selected site
certainly won't have a substantially wider cline than the selected cline.
\plr{Figure showing this? 
    For instance, either: (a) time-genome heatmap of number of heterozygotes (from PDE);
    or (b) a few time-space heatmaps of frequencies at different values of r (from PDE);
    or (c) a few space-genome heatmaps of frequencies at different times (from sims)
}
Of course, this calculation assumes that every generation provides an opportunity for recombination 
onto the other background,
but only recombinations in heterozygotes will move the linked lineage between selected backgrounds.
Since heterozygotes for the selected allele are only found at high frequency in the cline,
and lineages are generally pushed away from the cline but have no bias far away,
the amount of time a lineage spends in heterozygotes should grow as $\sqrt{\tau}$ for large $\tau$,
and so the width of the genomic region showing clines about the selected locus could be 
substantially larger than $1/\tau$.
However, this distinction appears hard to observe for realistic parameter values. \alisa{perhaps move some of this to discussion?}

@@

Most attempts to identify loci under selection in hybrid zones do so by identifying exceptional allele frequencies as a function of geographic position or genome-wide admixture proportion \citep{Gompert2012,OTHERS}.  
Current genomic approaches \alisa{applied} to the study of hybrid zones provide a additional rich sources of information -- namely ancestry deconvolution facilitated by programs such as hapMIX, LAMP and fineSTRUCTURE \yb{(CITE)} provides researchers with a measure of the length of ancestry tracts across the genome.  \alisa{In particular,} the information contained in the length distribution of these ancestry tracts genome-wide  has been extensively used to infer the demographic history of hybridization and admixture.  \alisa{However,the impact of selection on ancestry tract lengths in hybrid zones has received much less attention. Here, we present an expectation of block-length patterns throughout a hybrid zone, and find that selection against hybrid incompatibilities leaves a distinct genomic signature in the form of long contiguous blocks of ancestry around these loci.}

@@

These two novel findings may be beyond detection in current research programs that often make use of low-density genetic markers for which phase is unknown. However, as sequencing prices continue to drop and long-phased reads become more common, access to data pertaining to these observations could further aid in identifying selected sites in hybrid zones.  \yb{(CITE doi:10.1038/nbt0515-433 )} %Startups use short-read data to expand long-read sequencing market Michael Eisenstein Nature Biotechnology 33, 433?435 (2015) doi:10.1038/nbt0515-433 Published online 12 May 2015
\alisa{In the meantime, patterns of elevated pairwise LD, which is more easily computed from readily available data, surrounding selected loci could offer an approximation to these findings.}

@@

\subsection{Assumptions}
We made numerous simplifying assumptions to derive our analytical and also in the execution of the simulations. %Thus before our results can be profitably used for inference, it is worthwhile to  reflect on our assumptions and their consequences. 
Foremost, we have assumed that selection acts against heterozygotes, and that there is a single underdominant locus. This situation is rare in nature, since underdominant loci cannot invade when rare, and thus this assumption is likely rarely met. \alisa{We believe, however, that this assumption is unlikely to drastically influence our results, and previous results demonstrating that underdominant loci shares similar properties as more realistic genetic models such as ecological selection, mean that our findings here are somewhat generalizable} \yb{(CITE: (BARTON and HEWITT 1989; BARTON and GALE 1993)}

The assumption about the density of selected loci, however, may have greater consequences. Our model, which focuses on a single selected site is likely to be still relevant to cases in which there are only a small number of selected loci scattered throughout the genome. However, our predictions may differ significantly from scenarios in which the density of selected sites is high. Specifically, numerous selected sites within one ancestry block would likely act to both  increase the strength of selection, and to generate a longer unit which cannot be effectively broken up by recombination   \yb{(CITE: (BARTON 1983; Kruuk 1989)}. Such regions would likely be surrounded by even greater chunks of unbroken ancestry -- a result reflecting  greater $s$ and relatively short map distance between linked incompatibilities -- \alisa{and will impact expectations of genome-wide patterns of clines and block lengths. In terms of detecting targets of selection, such sites will be a common limitation among many approaches to identifying selection.}

For ease of calculations, our analytic model assumes a continuous and smooth one-dimensional spatial structure of the geographic landscape.

Lastly, the analytic model presented here lacks coalescence. \alisa{This can result in an underestimate of covariance in ancestry, as we have ignored sharing through genealogy.} The broad agreement between our analytic predictions and simulations suggests that this assumption does not impact our general results.

@@

Another important assumption is the smooth and one dimensional spatial structure of the geographic landscape.  
Patchily connected populations, or other esoteric forms of population structure may predominate on the edges of species ranges' where hybrid zones may occur.  
While, it is still likely that the genomic signature of longer chunks of ancestry around selected sites will hold in these situations, the strength of this expectation is unclear.

The tracking lineages given selection trick is an example of the more general idea of duality for particle systems.
It also assumes a large pop density limit that excludes recombination and makes lineages Markov;
if things are very patchy this might not be a good one.
Also, parent-offspring dispersal is not the same thing as successful-offspring-parent dispersal.

\alisa{epistasis}

Topics include\\
- ignoring coalescent\\
- single locus model, selection is STRONG\\

\subsection{Islands of divergence}  
The past years have witnessed evocative metaphors f seascapes to describe heterogeneous patterns of genetic divergence across the genome.  
The metaphors of "islands", "continents" and "archipelagoes" of divergence have inspired much research (and controversy) speciation genomics, but have been largely descriptive.  
We hope that our work provides a path towards thinking about the process of heterogeneous divergence more concretely. % in terms of the processes of migration, and selection influencing patterns of ancestry across the genome of hybridizing populations.
Specifically, we provide a natural interpretation of large contiguous chunks of ancestry, as sites linked to those experiencing selection.   
Overall focusing on lengths of ancestry blocks across the genome brings focus onto the processes of migration and selection rather than high-level summaries far severely abstracted from the evolutionary process.

>>>>

Within a genome, the mixing of ancestries is brought about by the the process of recombination that takes place every generation \citep{Fisher1954, Chapman2002, Baird2003}.   
The expected distribution of lengths of unbroken ancestry have been previously described under neutral conditions \cite[e.g.][]{Gravel2012,Sedghifar2015}.  
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection. 
Selection against migrant \yb{alleles} in a hybrid zone results in reduced frequencies of foreign ancestry surrounding targets of selection \yb{(CITE CLINES)}. 
The effect on patterns of ancestral block lengths is, however, less well understood. 
Because of rapid removal from the population, we expect to observe \yb{a deficit of} short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. 
\alisa{While empirical patterns of block length in hybrid zones have not been well studied, the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!

@@

\section{Methods}
\subsection*{The Model}
We consider two isolated populations --- labeled \emph{Species $A$} and \emph{Species $B$} ---  that come into contact $\tau$ generations in the past. 
Individuals move forward in time by sampling a displacement from a Gaussian distribution. 
We consider a simple case of under-dominance. 
\yb{Although most selection in hybrid zones is likely more complex than the single locus intrinsic under-dominance we consider, previous work has shown that such models generate clines that are effectively indistinguishable from more complex models of ecological selection \citep{Barton1989, Barton1993} which also generate clines like those degenerated under complex epistatic systems}. Specifically, here the fitness of an individual is  determined by its genotype at a particular locus $L$, such that individuals that are heterozygous at locus $L$ have fitness reduced by $s>0$ compared to homozygotes. To fix notation, we will say that species $A$ was initially on the ``left'' of the zone of contact, which corresponds to spatial positions $x<0$.

@@

\paragraph{Establishment of  \sout{the selected cline} \alisa{a cline at the selected locus}}
Consider a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline.
To study gene flow past this cline we first need to understand how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring.
Assuming local random assortment of alleles into diploids,
that the habitat and dispersal are homogeneous,
and that $s$ and $\sigma$ are small, 
the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{Bazykin1969}; for discussion of diploidy see \citep{diploidcline} \alisa{not sure which reference this is}.
There is not an exact analytic solution to this equation, 
but a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}.
Considering the numerous approximations going into this,
the important attributes are that 
the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially.
Furthermore, the cline is established through diffusion \alisa{of individuals?}, which is slowed down somewhat by selection against heterozygotes,
but is still diffusive, and so it will take time of order $1/s$
to reach its stable width \alisa{(is this something that has been shown before?)} This can be shown by rescaling space by $\sigma/s$ and time by $1/s$ in equation \eqref{eqn:cline_pde}, and seeing that the result \alisa{(what result?)} doesn't depend on $\sigma$ or on $s$.

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is given.
We will understand how lineages of sampled individuals behave,
both at and away from the selected site,
by following lineages as they move between backgrounds,
as done for instance in \citet{selectioncoal,Ralph2015}.
Consider the collection of $A$ alleles found at location $x$.
The expected number of such alleles produced by location $y$ is proportional to 
the number of $A$ alleles at $y$ multiplied by the probability of dispersing from $y$ to $x$;
looking in the other direction, this is also the probability that an $A$ allele at $x$ had a parent at $y$.
In other words, lineages move as the random walk determined by dispersal,
but biased towards areas that produce more of the selected allele that they carry.
Making the same assumptions underlying equation \eqref{eqn:cline_pde},
we show in Appendix \ref{apx:lineage_derivation}
that the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $s \grad \log p(x,t)$ (i.e., it is pushed uphill on $\log p$ at speed $s$) \alisa{note to self to read through appendix when written}.
Since $p=1$ far from the zone to the left, 
and is proportional to $\exp(-x\sqrt{s}/\sigma)$ to the right,
then roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma/\sqrt{s}$ 
back towards the side where they are more common.
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact,
this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction.

@@

\paragraph{How fast do linked clines relax?}
Although loci not under selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{Barton1986}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
Suppose we sample across at cline $\tau$ units of time after secondary contact.
How far away from the selected locus do we expect to see clines that have begun to flatten out more
than the selected cline?
A linked cline will have begun to flatten out if there is an appreciable chance that
a lineage at the linked locus initially linked to an $A$ allele 
at some point has recombined from an individual carrying the $B$ allele.
This implies that loci closer than $1/\tau$ to the selected site
certainly won't have a substantially wider cline than the selected cline.
\plr{Figure showing this? 
    For instance, either: (a) time-genome heatmap of number of heterozygotes (from PDE);
    or (b) a few time-space heatmaps of frequencies at different values of r (from PDE);
    or (c) a few space-genome heatmaps of frequencies at different times (from sims)
}
Of course, this calculation assumes that every generation provides an opportunity for recombination 
onto the other background,
but only recombinations in heterozygotes will move the linked lineage between selected backgrounds.
Since heterozygotes for the selected allele are only found at high frequency in the cline,
and lineages are generally pushed away from the cline but have no bias far away,
the amount of time a lineage spends in heterozygotes should grow as $\sqrt{\tau}$ for large $\tau$,
and so the width of the genomic region showing clines about the selected locus could be 
substantially larger than $1/\tau$.
However, this distinction appears hard to observe for realistic parameter values. \alisa{perhaps move some of this to discussion?}

@@

Most attempts to identify loci under selection in hybrid zones do so by identifying exceptional allele frequencies as a function of geographic position or genome-wide admixture proportion \citep{Gompert2012,OTHERS}.  
Current genomic approaches \alisa{applied} to the study of hybrid zones provide a additional rich sources of information -- namely ancestry deconvolution facilitated by programs such as hapMIX \citep{Price2009}, LAMP \citep{Sankararaman2008} and fineSTRUCTURE \citep{Lawson2012} provides researchers with a measure of the length of ancestry tracts across the genome.  \alisa{In particular,} the information contained in the length distribution of these ancestry tracts genome-wide  has been extensively used to infer the demographic history of hybridization and admixture.  \alisa{However,the impact of selection on ancestry tract lengths in hybrid zones has received much less attention. Here, we present an expectation of block-length patterns throughout a hybrid zone, and find that selection against hybrid incompatibilities leaves a distinct genomic signature in the form of long contiguous blocks of ancestry around these loci.}

@@

These two novel findings may be beyond detection in current research programs that often make use of low-density genetic markers for which phase is unknown. However, as sequencing prices continue to drop and long-phased reads become more common, access to data pertaining to these observations could further aid in identifying selected sites in hybrid zones.  %\yb{(CITE doi:10.1038/nbt0515-433 )} %Startups use short-read data to expand long-read sequencing market Michael Eisenstein Nature Biotechnology 33, 433?435 (2015) doi:10.1038/nbt0515-433 Published online 12 May 2015
\alisa{In the meantime, patterns of elevated pairwise LD, which is more easily computed from readily available data, surrounding selected loci could offer an approximation to these findings.}

@@

\subsection{Assumptions}
We made numerous simplifying assumptions to derive our analytical and also in the execution of the simulations. %Thus before our results can be profitably used for inference, it is worthwhile to  reflect on our assumptions and their consequences. 
Foremost, we have assumed that selection acts against heterozygotes, and that there is a single underdominant locus. This situation is rare in nature, since underdominant loci cannot invade when rare, and thus this assumption is likely rarely met. \alisa{We believe, however, that this assumption is unlikely to drastically influence our results, and previous results demonstrating that underdominant loci shares similar properties as more realistic genetic models such as ecological selection, mean that our findings here are somewhat generalizable} \citep{Barton1989; Barton1993}

The assumption about the density of selected loci, however, may have greater consequences. Our model, which focuses on a single selected site is likely to be still relevant to cases in which there are only a small number of selected loci scattered throughout the genome. However, our predictions may differ significantly from scenarios in which the density of selected sites is high. Specifically, numerous selected sites within one ancestry block would likely act to both  increase the strength of selection, and to generate a longer unit which cannot be effectively broken up by recombination   \citep{Barton1983; Kruuk1999)}. Such regions would likely be surrounded by even greater chunks of unbroken ancestry -- a result reflecting  greater $s$ and relatively short map distance between linked incompatibilities -- \alisa{and will impact expectations of genome-wide patterns of clines and block lengths. In terms of detecting targets of selection, such sites will be a common limitation among many approaches to identifying selection.}

For ease of calculations, our analytic model assumes a continuous and smooth one-dimensional spatial structure of the geographic landscape. This allows us to derive analytic expressions that track lineages, and is an example of the more general idea of duality for particle systems \alisa{I think this may be beyond the average ME reader?}. However, this assumes a large population density limit that excludes recombination and makes linages Markov \alisa{not sure about what point this is making? Is there a better way to state this for our intended audience?}.  On the other hand, our simulations model regularly spaced demes. In reality, populations may be patchily connected, especially at the edges of species ranges where hybrid zones may occur. While we still believe that the genomic signature of long ancestry chunks around selected sites will still hold in these situations, the strength of this expectation is unclear.\\

Lastly, the analytic model presented here lacks coalescence. \alisa{This can result in an underestimate of covariance in ancestry, as we have ignored sharing through genealogy \cite{Liang2014}.} The broad agreement between our analytic predictions and simulations suggests that this assumption does not impact our general results.

@@

Also, parent-offspring dispersal is not the same thing as successful-offspring-parent dispersal.

\alisa{epistasis}
\alisa{selection here is STRONG, not sure how well this will work on systems in which selection on any given locus is small (most systems)}

\subsection{Islands of divergence}  
The past years have witnessed evocative metaphors of seascapes to describe heterogeneous patterns of genetic divergence across the genome.  
The metaphors of "islands", "continents" and "archipelagoes" of divergence have inspired much research (and controversy) speciation genomics, but have been largely descriptive.  
We hope that our work provides a path towards thinking about the process of heterogeneous divergence more concretely. % in terms of the processes of migration, and selection influencing patterns of ancestry across the genome of hybridizing populations.
Specifically, we provide a natural interpretation of large contiguous chunks of ancestry, as sites linked to those experiencing selection.   
Overall focusing on lengths of ancestry blocks across the genome brings focus onto the processes of migration and selection rather than high-level summaries far severely abstracted from the evolutionary process.
